---
title: "HW3: Distributional consequences of climate policy"
author: "Drew Andersen"
date: "5/12/2022"
output: 
  html_document:
    code_folding: hide
---
```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(here)
library(tidyverse)
library(kableExtra)
```
# Introduction

Use the following set of facts:

• Consumers can be separated into two income groups: “high” and “low.” The data set provides price (in $) and quantity (in kWh) estimates of demand per month for the two groups. Run linear regressions (with an intercept) to estimate the demand curves for “high” and “low” income consumers.

• Initially, there is no tax on electricity consumption.

• The current electricity price (without any taxes) is $.10 per kWh.

• The marginal cost of producing a kWh of electricity is linear and has a price-intercept of 0.

```{r}
#Loading in the data and baselines

data <- read_csv(here("HW3_data.csv"))[,-1] #upload and drop first column

#est constants
price_e = 10 #cents/kWh
scc = 5100 # cent price/metric_ton 1000 kg
emis_lb_per_kwh = .85 #lbs CO2/kWh
emis_kg_per_kwh = emis_lb_per_kwh * 0.45359237 #kgs CO2/kWh
```

```{r}
#linear regression for functions
low_lm <- lm(Price_Cents ~ Q_Low_Kwh, data = data)
high_lm <- lm(Price_Cents ~ Q_High_Kwh, data = data)

agg_slope <- 1/((1/low_lm$coefficients[2])+(1/high_lm$coefficients[2]))

high_dem <- function(quantity){  
  demand_p = high_lm$coefficients[2]*quantity + high_lm$coefficients[1]
  return(demand_p)
}
low_dem <- function(quantity){  
  demand_p = low_lm$coefficients[2]*quantity + low_lm$coefficients[1]
}
agg_dem <- function(quantity){
  demand_p = agg_slope*quantity + high_lm$coefficients[1]
  return(demand_p)
}

#linear regression for inverse functions 
low_lm_inv<- lm(Q_Low_Kwh ~ Price_Cents, data = data)
high_lm_inv <- lm(Q_High_Kwh ~ Price_Cents, data = data)

high_dem_inv <- function(price){  
  demand_q = high_lm_inv$coefficients[2]*price + high_lm_inv$coefficients[1]
}
low_dem_inv <- function(price){  
  demand_q = low_lm_inv$coefficients[2]*price + low_lm_inv$coefficients[1]
}
agg_dem_inv <- function(price){  
  demand_q = (price - high_lm$coefficients[1])/agg_slope
}
```

##Q1
```{r}
mec = emis_kg_per_kwh * scc * (1/1000) #mec per kWh electricity
```
The marginal externality cost per kwH of electricity is $`r round(mec,2)`cents/kwH.

##Q2
```{r}
quant_e <- agg_dem_inv(price_e)
supply_slope <- price_e/quant_e

sup <- function(quantity){
  price = supply_slope * quantity
}
sup_inv <- function(price){
  quantity = price/supply_slope
}
```

```{r, fig.align="center"}
ggplot(data = data, aes(x = Q_Kwh, y = Price_Cents)) +
  geom_abline(intercept = high_lm$coefficients[1], slope = agg_slope, color = "black") +
  geom_abline(intercept = 0, slope = supply_slope, color = "purple") + 
  geom_abline(intercept = high_lm$coefficients[1], slope = high_lm$coefficients[2], color = "green") +
  geom_abline(intercept = low_lm$coefficients[1], slope = low_lm$coefficients[2], color = "blue") +
  theme_minimal() + 
  ylim(0,35) +
  xlim(0,800000)
```

**Figure 1:** The demand curves for both the "Low" & "High" Groups are plotted above in blue and green, respectively. The black line represents the aggregate demand. The purple line represents the supply curve. 

```{r}
con_surplus = (.5*quant_e*(high_lm$coefficients[1] - price_e))/100
prod_surplus = (.5*quant_e*price_e)/100
env_cost = (mec * quant_e)/100 #div by hundred for dollar value
```
*This model yields over estimates because the kink in the aggregate demand curve was not captured in this model. 

The benefit to consumers is $`r format(round(con_surplus,0), scientific=F)` 

The benefit to producers is $`r format(round(prod_surplus,0), scientific=F)` 

The environmental cost is $`r format(round(env_cost,0),scientific=F)`

##Q3
```{r}
high_ben = (.5 * high_dem_inv(price_e) * (high_lm$coefficients[1] - price_e)) / 100
low_ben = (.5 * low_dem_inv(price_e) * (low_lm$coefficients[1] - price_e)) / 100
```

The consumer benefit for the "High" group is $`r format(round(high_ben,0), scientific=F)`

The consumer benefit for the "Low" group is $`r round(low_ben,0)`

##Q4
```{r}
new_q = (mec - high_lm$coefficients[1])/(agg_slope-supply_slope)
new_p = agg_dem(new_q)

high_ben_tax = (.5 * high_dem_inv(new_p) * (high_lm$coefficients[1] - new_p)) / 100
low_ben_tax = (.5 * low_dem_inv(new_p) * (low_lm$coefficients[1] - new_p)) / 100

prod_surplus_tax = (.5*new_q*(new_p-mec))/100
env_cost_tax = (mec * new_q)/100
tax_rev = new_q * (mec)/100
```
(a) Quantity (kwh) = `r format(round(new_q,0),scientific=F)` kwh
(b) Price (cents) = `r round(new_p,1)` cents
(c) "High" Group Welfare = $`r format(round(high_ben_tax,0),scientific=F)`
(d) "Low" Group Welfare = $`r round(low_ben_tax,0)`
(e) Power Supplier Surplus = $`r format(round(prod_surplus_tax,0),scientific=F)`
(f) Environmental Cost = $`r format(round(env_cost_tax,0),scientific=F)`
(g) Total tax revenue = $`r format(round(tax_rev,0),scientific=F)`

##Q5
```{r}
scc = 5100
mec = emis_kg_per_kwh * scc * (1/1000) #mec per kWh electricity

new_q = (mec - high_lm$coefficients[1]) / (agg_slope - supply_slope)
new_p = agg_dem(new_q)

tax_rev = new_q * mec/100

high_perc = high_dem_inv(new_p)/new_q
low_perc = 1 - high_perc

high_tax_share = tax_rev * high_perc
low_tax_share = tax_rev * low_perc

high_ben_plus_tax = ((.5 * high_dem_inv(new_p) * (high_lm$coefficients[1] - new_p)) / 100) + high_tax_share
low_ben_plus_tax = ((.5 * low_dem_inv(new_p) * (low_lm$coefficients[1] - new_p)) / 100) + low_tax_share
prod_surplus_tax = (.5*new_q*(new_p-mec))/100
```

SCC = $51

(a) "High" Income Consumer Benefit = $`r format(round(high_ben_plus_tax,0), scientific=F)`
(b) "Low" Income Consumer Benefit = $`r format(round(low_ben_plus_tax,0), scientific=F)`
(c) Producer Benefit = $`r format(round(prod_surplus_tax,0), scientific=F)`

```{r}
scc = 7500
mec = emis_kg_per_kwh * scc * (1/1000) #mec per kWh electricity

new_q = (mec - high_lm$coefficients[1]) / (agg_slope - supply_slope)
new_p = agg_dem(new_q)

tax_rev = new_q * mec/100

high_perc = high_dem_inv(new_p)/new_q
low_perc = 1 - high_perc

high_tax_share = tax_rev * high_perc
low_tax_share = tax_rev * low_perc

high_ben_plus_tax = ((.5 * high_dem_inv(new_p) * (high_lm$coefficients[1] - new_p)) / 100) + high_tax_share
low_ben_plus_tax = ((.5 * low_dem_inv(new_p) * (low_lm$coefficients[1] - new_p)) / 100) + low_tax_share
prod_surplus_tax = (.5*new_q*(new_p-mec))/100
```

SCC = $75

(a) "High" Income Consumer Benefit = $`r format(round(high_ben_plus_tax,0), scientific=F)`
(b) "Low" Income Consumer Benefit = $`r format(round(low_ben_plus_tax,0), scientific=F)`
(c) Producer Benefit = $`r format(round(prod_surplus_tax,0), scientific=F)`

```{r}
scc = 10000
mec = emis_kg_per_kwh * scc * (1/1000) #mec per kWh electricity

new_q = (mec - high_lm$coefficients[1]) / (agg_slope - supply_slope)
new_p = agg_dem(new_q)

tax_rev = new_q * mec/100

high_perc = high_dem_inv(new_p)/new_q
low_perc = 1 - high_perc

high_tax_share = tax_rev * high_perc
low_tax_share = tax_rev * low_perc

high_ben_plus_tax = ((.5 * high_dem_inv(new_p) * (high_lm$coefficients[1] - new_p)) / 100) + high_tax_share
low_ben_plus_tax = ((.5 * low_dem_inv(new_p) * (low_lm$coefficients[1] - new_p)) / 100) + low_tax_share
prod_surplus_tax = (.5*new_q*(new_p-mec))/100
```

SCC = $100

(a) "High" Income Consumer Benefit = $`r format(round(high_ben_plus_tax,0), scientific=F)`
(b) "Low" Income Consumer Benefit = $`r format(round(low_ben_plus_tax,0), scientific=F)`
(c) Producer Benefit = $`r format(round(prod_surplus_tax,0), scientific=F)`


```{r}
scc = 15000
mec = emis_kg_per_kwh * scc * (1/1000) #mec per kWh electricity

new_q = (mec - high_lm$coefficients[1]) / (agg_slope - supply_slope)
new_p = agg_dem(new_q)

tax_rev = new_q * mec/100

high_perc = high_dem_inv(new_p)/new_q
low_perc = 1 - high_perc

high_tax_share = tax_rev * high_perc
low_tax_share = tax_rev * low_perc

high_ben_plus_tax = ((.5 * high_dem_inv(new_p) * (high_lm$coefficients[1] - new_p)) / 100) + high_tax_share
low_ben_plus_tax = ((.5 * low_dem_inv(new_p) * (low_lm$coefficients[1] - new_p)) / 100) + low_tax_share
prod_surplus_tax = (.5*new_q*(new_p-mec))/100
```

SCC = $125

(a) "High" Income Consumer Benefit = $`r format(round(high_ben_plus_tax,0), scientific=F)`
(b) "Low" Income Consumer Benefit = $`r format(round(low_ben_plus_tax,0), scientific=F)`
(c) Producer Benefit = $`r format(round(prod_surplus_tax,0), scientific=F)`


```{r}
scc = 12500
mec = emis_kg_per_kwh * scc * (1/1000) #mec per kWh electricity

new_q = (mec - high_lm$coefficients[1]) / (agg_slope - supply_slope)
new_p = agg_dem(new_q)

tax_rev = new_q * mec/100

high_perc = high_dem_inv(new_p)/new_q
low_perc = 1 - high_perc

high_tax_share = tax_rev * high_perc
low_tax_share = tax_rev * low_perc

high_ben_plus_tax = ((.5 * high_dem_inv(new_p) * (high_lm$coefficients[1] - new_p)) / 100) + high_tax_share
low_ben_plus_tax = ((.5 * low_dem_inv(new_p) * (low_lm$coefficients[1] - new_p)) / 100) + low_tax_share
prod_surplus_tax = (.5*new_q*(new_p-mec))/100
```

SCC = $150

(a) "High" Income Consumer Benefit = $`r format(round(high_ben_plus_tax,0), scientific=F)`
(b) "Low" Income Consumer Benefit = $`r format(round(low_ben_plus_tax,0), scientific=F)`
(c) Producer Benefit = $`r format(round(prod_surplus_tax,0), scientific=F)`

##Q6
```{r}
scc = 5100
mec = emis_kg_per_kwh * scc * (1/1000) #mec per kWh electricity

data_solar <- data %>% 
  mutate(Solar_Price = Price_Cents/2)

solar_lm <- lm(Solar_Price ~ Q_High_Kwh, data = data_solar)

solar_agg_slope <- 1/(low_lm_inv$coefficients[2] + high_lm_inv$coefficients[2])

solar_demand <- function(quantity){  
  demand_p = solar_agg_slope[2]*quantity + low_lm$coefficients[1]
  return(demand_p)
}

new_q <- (low_lm$coefficients[1]) / (supply_slope - solar_agg_slope)
new_p <- supply_slope*new_q

env_cost_solar = (mec * new_q)/100

tax = env_cost_solar*100/(new_q)
```

(a) Total electricity consumption: `r format(round(new_q,0), scientific=F)` kwh
(b) Total environmental externality: $`r round(env_cost_solar,0)`
(c) Equivalent tax: $`r round(tax,2)` cents/kwh
